---
title: "Trees"
author: "Moritz Baldauf, Christian Vorhauser"
format: html
editor: visual
---

## 

## Predicting how a driver would perform if put into a different Team

### Using Tree-based methods

```{r, warning=FALSE}
library(rpart)
library(party)
library(ranger)
library(partykit)
```

```{r}

set.seed(134)

F1_complete <- read.csv("data/F1_preprocessed",stringsAsFactors = T)
F1_complete$X <- NULL
F1 <- F1_complete[,1:12] # Remove points because of high correlation to Finishing Position
head(F1)
tail(F1)
str(F1)

summary(F1)

```

```{r}
tree <- rpart(FinishingPosition~., data = F1)
plot(as.party(tree))
# We see that points is the biggest inpacting factor on starting postions
# No suprise as you get a lot of points if you finish in a good postion 
```

```{r}
ncol(F1)
start_time <- Sys.time()
tree2 <- rpart(FinishingPosition~., data = F1)
end_time <- Sys.time()
execution_time <- end_time - start_time
print(execution_time)
```

```{r}
plot(as.party(tree2))

full_tree <- rpart(FinishingPosition~., data = F1, control = list(cp = 0))
rpart::plotcp(full_tree) # Super hard to read anything

full_tree_zoom <- rpart(FinishingPosition~., data = F1, control = list(cp = 0.005)) # Zoom in
rpart::plotcp(full_tree_zoom)
printcp(full_tree_zoom)

```

```{r}

pruned_tree <- prune(full_tree, cp = full_tree$cptable[5, "CP"])

plot(as.party(pruned_tree))

```

```{r}

yhat <- predict(pruned_tree,data = F1)
hist(yhat, main = "Title",breaks = 24) # Not exactly sure what this means, but it kinda makes sens that we have more observations on the higher end because of people not finishing


```

```{r}

#predict(pruned_tree, )

#str(F1)
raceId <- 1102
year <- 2024
round <- 1 
country <- "bahrain"
officialName <- "Formula 1 Gulf Air Bahrain Grand Prix 2024"
driverId <- "max-verstappen"
constructorId <- "red-bull"
engineManufacturerId <- "honda-rbpt"
PositionPractice1 <- 1
PositionPractice2 <- 4
StartingPosition <- 2

colnames(F1)
#str(F1)
Predictor_df <- data.frame(raceId, year, round, country, officialName, driverId, constructorId, engineManufacturerId, PositionPractice1, PositionPractice2, StartingPosition)

predict(pruned_tree, Predictor_df)

```

## RF

```{r}
set.seed(345)
F1[is.na(F1$StartingPosition),]$StartingPosition <- 0  
rf <- ranger(FinishingPosition ~. , data = F1, probability = FALSE, importance = "permutation")

print(rf)

plot(as.table(importance(rf)))
```

From the importance plot we see that, surprisingly, round, country and driverId are the least important ones. The positions in the Practices and Qualifying are the most important, as they influence the starting position and winning from the front is generally easier.

### Partial dependency plot

```{r}
grid <- seq(1,24,length.out = 24)
nd <- F1[sample.int(nrow(F1),1000),]
prs <- lapply(grid, \(val) {
  nd$StartingPosition <- val
  predict(rf, data = nd, type = "response")$predictions
})

matplot(grid, t(do.call("cbind",prs)), type = "l",lty = 1, col = rgb(.1,.1,.1,.1))
```

Here we can see that the higher the StartingPosition the more important it is in splitting the trees, while for the first 10 Starting positions it is not too important.

## Predicting Finishing Position for 2 drivers

We use Max Verstappen and Logan Sargeant for this example and also switch their cars to see the influence the model predicts. The race will be Bahrain 2024.

```{r}
set.seed(345)

raceId <- 1102
year <- 2024
round <- 1 
country <- "bahrain"
officialName <- "Formula 1 Gulf Air Bahrain Grand Prix 2024"
driverId <- "max-verstappen"
constructorId <- "red-bull"
engineManufacturerId <- "honda-rbpt"
PositionPractice1 <- 6
PositionPractice2 <- 6
StartingPosition <- 1

nd <- data.frame(raceId, year, round,country,officialName, driverId, constructorId, engineManufacturerId, PositionPractice1, PositionPractice2, StartingPosition)

predict(rf, data = nd, type = "response")$predictions

raceId <- 1102
year <- 2024
round <- 1 
country <- "bahrain"
officialName <- "Formula 1 Gulf Air Bahrain Grand Prix 2024"
driverId <- "logan-sargeant"
constructorId <- "red-bull"
engineManufacturerId <- "honda-rbpt"
PositionPractice1 <- 16
PositionPractice2 <- 13
StartingPosition <- 18

nd <- data.frame(raceId, year, round,country,officialName, driverId, constructorId, engineManufacturerId, PositionPractice1, PositionPractice2, StartingPosition)

predict(rf, data = nd, type = "response")$predictions

raceId <- 1102
year <- 2024
round <- 1 
country <- "bahrain"
officialName <- "Formula 1 Gulf Air Bahrain Grand Prix 2024"
driverId <- "max-verstappen"
constructorId <- "williams"
engineManufacturerId <- "mercedes"
PositionPractice1 <- 6
PositionPractice2 <- 6
StartingPosition <- 1

nd <- data.frame(raceId, year, round,country,officialName, driverId, constructorId, engineManufacturerId, PositionPractice1, PositionPractice2, StartingPosition)

predict(rf, data = nd, type = "response")$predictions

raceId <- 1102
year <- 2024
round <- 1 
country <- "bahrain"
officialName <- "Formula 1 Gulf Air Bahrain Grand Prix 2024"
driverId <- "logan-sargeant"
constructorId <- "williams"
engineManufacturerId <- "mercedes"
PositionPractice1 <- 16
PositionPractice2 <- 13
StartingPosition <- 18

nd <- data.frame(raceId, year, round,country,officialName, driverId, constructorId, engineManufacturerId, PositionPractice1, PositionPractice2, StartingPosition)

predict(rf, data = nd, type = "response")$predictions
```

We see that the model predicts base-Max to finish in 4th place given his parameters and 8th for driving in Logans car. Vice versa, base-Logan is predicted to finish in 18th place while Logan in the RedBull is predicted to finish in 16th place.

From this we conclude that the car does make a big difference in the finishing Position of the race.

## Clustering😩

PCA only works for numeric, especially continuous values, so we cannot do it (sad)

```{r}
# random interesting graphs maybe
plot(F1$constructorId,F1$FinishingPosition)
plot(F1$driverId,F1$FinishingPosition)
plot(F1$raceId,F1$engineManufacturerId)
```

We create a subset of the dataframe which groups the data by constructor and year and sums their points. We also only use 2010 to 2023 as the scoring system changed in 2010 and 2024 has too few races.

```{r}
# creating datapoints for clustering, points scored per team per year
# we cluster into high scorers, mid-scorers and low scorers
library(dplyr)

pts_const_y <- F1_complete %>%
                  group_by(constructorId,year) %>%
                  summarize(sum(points))
colnames(pts_const_y) <- c("constructorId","year","points")
pts_const_y <- pts_const_y[which(pts_const_y$year >= 2010 & pts_const_y$year < 2024 ),]
plot(pts_const_y$year,pts_const_y$points,col = pts_const_y$constructorId)
```

We make 4 clusters which can be interpreted as "very little points", "did alright", "did great", "top team(s)".

```{r}
set.seed(1)
kms <- kmeans(pts_const_y[,2:3], centers = 4, nstart = 10)
plot(pts_const_y$year,pts_const_y$points, col = kms$cluster+1,pch = 18)
text(pts_const_y$year,pts_const_y$points, labels = pts_const_y$constructorId, adj = c(-0.15,0.5),cex = 0.7)
```

To give an overview on how often the teams were in the clusters. Cluster numbering is as follows: 1 - "top team(s)", 4 - "did great", 2 - "did alright", 3 - "very little points".

```{r}
pts_const_y$cluster <- kms$cluster
clusters_overview <- pts_const_y %>%
                        group_by(cluster,constructorId) %>%
                        summarise(Count = n())
clusters_overview

clusters_overview %>%
  group_by(cluster) %>%
  summarise(numTeams = n())
```
